---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Data Summaries {#Start3}

```{r tidyr3, include = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

Let's create some data summaries to see what is going on with the Nightjar survey program.

First, import the data and reload your libraries if you are starting a new session. 
```{r data lib, message=FALSE, eval=FALSE}

require(tidyverse)
require(lubridate)
require(data.table)
require(reshape2)
require(tibble)

dat<-read.csv("Data/Filter_Nightjar.csv")

out.dir <- paste("Output/")
dat.dir <- paste("Data/")

```

Look at the number of stops per route in each year. Keep in mind that anything between 6 - 12 stops per route is normal. We therefore append the number of stop to the original dataframe in order to allow for effort correction later in the analysis  
```{r nstop, eval=FALSE, message=FALSE}

nstop<-dat %>% group_by(survey_year, SiteCode) %>% summarize(nstop=n_distinct(SurveyAreaIdentifier))

dat<-left_join(dat, nstop, by=c("survey_year", "SiteCode"))

#Useful table for the coordinator to check that the data are entered properly.
nstop1<-cast(nstop, SiteCode~survey_year, value = "nstop")
write.csv(nstop1, "Output/Nigthjar.nstop.csv")

write.csv(dat, "Data/Filter_Nightjar.csv")

```

Create an survey events layer for when we need to zero-fill the data matrix. 
```{r zero, eval= FALES, message=FALSE}

event.data <- dat %>%
select(SamplingEventIdentifier, SurveyAreaIdentifier, SiteCode, survey_year, CollectorNumber, Prov, nstop) %>%  
distinct(SamplingEventIdentifier, .keep_all=TRUE) %>%
as.data.frame()

write.csv(event.data, "Output/Events.csv")
```

Count the number of routes per years and province
```{r nroute, message=FALSE, eval=FALSE}
nroute<-dat %>% group_by(survey_year) %>% summarize(nroute=n_distinct(SiteCode))

nroute_prov<-dat %>% group_by(survey_year, Prov) %>% summarize(nroute_prov= n_distinct(SiteCode))

nroute_prov1<-cast(nroute_prov, survey_year~Prov, value="nroute_prov")
write.csv(nroute_prov1, "Output/Nigthjar.nroute.prov.csv")

plot_route<-ggplot(nroute_prov, aes(survey_year, nroute_prov))+
  geom_line(aes(colour=Prov), size=1)+
  theme_classic()+
  theme(text=element_text(size=20))+
  ylab("Number of Routes")

```

Count the number of birds detected
```{r nbirds, message=FALSE, eval=FALSE}

ncount<-dat %>% filter(ObservationCount==1) %>% droplevels()
ncount<-ncount %>% filter(CommonName %in% c("Common Nighthawk", "Common Poorwill", "Eastern Whip-poor-will"))
ncount<-ncount %>% group_by(survey_year, CommonName) %>% summarise(ncount=sum(ObservationCount))

ncount<-cast(ncount, survey_year~CommonName, value="ncount")
write.csv(ncount, "Output/Nighjar.ncount.csv")

```

Mean count per route, for three target species

```{r mean}

r_mean<-dat %>% filter(ObservationCount==1) %>% droplevels()
r_mean<-r_mean %>% filter(CommonName %in% c("Common Nighthawk", "Common Poorwill", "Eastern Whip-poor-will"))
r_mean<-r_mean %>% group_by(survey_year, Prov, CommonName) %>% summarise(r_tot=sum(ObservationCount))

r_mean<-left_join(r_mean, nroute_prov, by=c("survey_year", "Prov"))

r_mean<-r_mean %>% mutate(r_mean=r_tot/nroute_prov)

##Plot is ugly
#ggplot(r_mean, aes(survey_year, r_mean))+
#  geom_line(aes(colour=Prov), size=1)+
#  facet_wrap(~CommonName, scales='free')+
#  theme_classic()+
#  theme(text=element_text(size=20))

```

Species specific processing notes: 
Any six-stop routes in the system at the moment from BC, AB and SK would be surveyed too early in the evening to pick up Common Poorwills, and that these routes should be removed for any analyses for this species, but they would still be okay for Common Nighthawks. First 6 stops best for Common Nighthawks and the second for either Common Poorwills or Eastern Whip-poor-wills


#now we will make a summary of the mean species per year. This needs to be a zero-filld data matrix. 
```{r speciesdat, echo = TRUE, message = FALSE}

#select desired data columns for analysis
obs<-dat %>% select("CommonName", "SamplingEventIdentifier", "SurveyAreaIdentifier", "SiteCode", "survey_year", "CollectorNumber", "Prov", "ObservationCount")

#select just target species
obs<-obs %>% filter(CommonName %in% c("Common Nighthawk", "Common Poorwill", "Eastern Whip-poor-will"))

```

We are ready to develop the species-specific analysis loop

```{r sp.loop, echo = TRUE, message = FALSE}

## Generate species list for analysis. The species that make the cut will vary on province based on the data manipulation

species.list <- unique(obs$CommonName)

# Create output table before entering the loop
mean_count<- as.data.frame(matrix(data = NA, nrow = 1, ncol = 4, byrow = FALSE, dimnames = NULL))

names(mean_count) <- c("Prov", "survey_year", "meancount", "Species")

write.table(mean_count, file = paste(out.dir,"Nightjar.meancount.csv", sep=""), row.names = FALSE, append = TRUE, col.names = TRUE, quote = FALSE, sep = ",")


## Create the loop
for(m in 1:length(species.list)) {

# for testing each species
# m<-3 
   
  sp.data <-NULL 
  sp.data <- filter(obs, CommonName == species.list[m]) %>%
      droplevels()
  
  print(paste("Currently analyzing species ", m, "/", species.list[m], sep = "")) 

##-----------------------------------------------------------

#zero fill by merging with the events data frame. 
sp.data <- left_join(event.data, sp.data, by = c("SamplingEventIdentifier", "SurveyAreaIdentifier",  "SiteCode", "survey_year", "CollectorNumber", "Prov")) %>% mutate(ObservationCount = replace(ObservationCount, is.na(ObservationCount), 0))

##----------------------------------------------------------

# Limit to species observed at least once per route 

# Summarize survey site to determine which species have been observed at least once (looking at the total count column) those with sum <= 1 across all survey years will be dropped from analysis (implies never observed on a route (i.e., outside range or inappropriate habitat))

  site.summ <- melt(sp.data, id.var = "SiteCode",	measure.var = "ObservationCount")
  site.summ$value<-as.numeric(site.summ$value)
  site.summ <- cast(site.summ, SiteCode ~ variable,	fun.aggregate="sum")
  site.sp.list <- unique(subset(site.summ, select = c("SiteCode"), ObservationCount >= 1))

# Limit raw data to these species, i.e., those that were observed at least once on a route 
  sp.data <- merge(sp.data, site.sp.list, by = c("SiteCode"))

##-----------------------------------------------------------

# Count the number of individuals per route as the response variable. The number of stop on a route will used as a covariate or offset in the model to control for route level effort.  

sp.data$ObservationCount<-as.numeric(sp.data$ObservationCount)
sp.data<-sp.data %>% group_by(SiteCode, survey_year, CollectorNumber, Prov, nstop) %>% summarise(count=sum(ObservationCount))

##Calculate mean count per province per year

sp.data<-as.data.frame(sp.data)

mean_count<-sp.data %>% group_by(Prov, survey_year) %>% summarize(mean_count=mean(count))
mean_count$Species<-species.list[m]

write.table(mean_count, file = paste(out.dir,"Nightjar.meancount.csv", sep=""), row.names = FALSE, append = TRUE, col.names = FALSE, quote = FALSE, sep = ",")

}#end species loop  

#now we can import the full data table and plot the results

mean_count<-read.csv("Output/Nightjar.meancount.csv")

```
