---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Occupancy Analysis {#Stat5}

```{r tidyr5, include = FALSE, warning = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)

```

Now we will create an occupancy model to see what is going on with the Nightjar survey program.

First, import the data and reload your libraries if you are starting a new session. 

```{r data lib, message=FALSE, eval=FALSE}

require(tidyverse)
require(lubridate)
require(reshape)
require(unmarked)
require(naturecounts)

dat<-read.csv("Data/Filter_Nightjar.csv")
event.data<-read.csv("Output/Events.csv")

out.dir <- paste("Output/")
dat.dir <- paste("Data/")

```

Now we want to prepare the data for the `unmarked` package which will accommodate our occupancy analysis. Since we have 6 X 1-min time intervals, we will use the repeated count model `pcount`. 

For an occupancy analysis, the input data will fall into 3 categories:

Detection histories in which the rows of the data will usually be sites and the columns will represent detection occasions or other units of replication.

Site-specific covariates, which are attributes measured on each of the sites that potentially explain site-to-site variability in detection and/or occupancy.

Sample-specific or "observation level" covariates, which are attributes measures at each of the sample occasions (replicates).

Detection Covariates = DOY, Wind speed, air temp, moon (Yes or No), hours since sunset.  

Abundance Covariates = Year, latitude, longitude 

#First we will create the detection history framework.
The matrix with observed detection history is 0's and 1's, one row per site, one column per survey.

Let's start by selecting the data that we need and creating any new fields that are essential to the `unmarked` framework. 

```{r clean data, eval=FALSA, message=FALSE}

#use the `naturecounts` helper function to create the date fields. 

dat<- format_dates(dat)

#Replace codes with numeric occupancy values
#W (wing-boom) = 1
#C (call) = 1
#V (visual) = 1
#N (not detected) = 1 
#U (unknown detection type) = 1

dat <- dat %>% mutate(V1=ifelse(ObservationCount2 %in% c("W", "C", "V", "U"), 1, 0), V2=ifelse(ObservationCount3 %in% c("W", "C", "V", "U"), 1, 0), V3=ifelse(ObservationCount4 %in% c("W", "C", "V", "U"), 1, 0), V4=ifelse(ObservationCount5 %in% c("W", "C", "V", "U"), 1, 0), V5=ifelse(ObservationCount6 %in% c("W", "C", "V", "U"), 1, 0), V6=ifelse(ObservationCount7 %in% c("W", "C", "V", "U"), 1, 0)) %>% 
select(-ObservationDescriptor, -ObservationCount, -ObservationDescriptor2, -ObservationCount2, -ObservationDescriptor3, -ObservationCount3, -ObservationDescriptor4, -ObservationCount4, -ObservationDescriptor5, -ObservationCount5, -ObservationDescriptor6, -ObservationCount6, -ObservationDescriptor7, -ObservationCount7,
-ObservationDescriptor8, -ObservationCount8,
-ObservationDescriptor9, -ObservationCount9, -X.1, -X, -Locality, -SamplingEventIdentifier)

#select just target species
dat<-dat %>% filter(CommonName %in% c("Common Nighthawk", "Common Poorwill", "Eastern Whip-poor-will"))

species.list <- unique(dat$CommonName)

## Create the speciesloop
for(m in 1:length(species.list)) {
  
  # for testing each species
m<-3 
   
  sp.data <-NULL 
  sp.data <- filter(dat, CommonName == species.list[m]) %>%
      droplevels()
  
 
  
} # end loop


```


